---
title: "Decision Tree Challenge"
subtitle: "Feature Importance and Categorical Variable Encoding"
format:
  html: default
  pdf: default
execute:
  echo: true
  eval: true
---

# ðŸŒ³ Decision Tree Challenge - Feature Importance and Variable Encoding

## Challenge Overview

---
title: "Decision Tree Challenge"
subtitle: "Feature Importance and Categorical Variable Encoding"
format:
  html: default
  pdf: default
execute:
  echo: true
  eval: true
---

# ðŸŒ³ Decision Tree Challenge - Feature Importance and Variable Encoding

## Challenge Overview

**Your Mission:** Create a comprehensive Quarto document that demonstrates how decision trees measure feature importance, analyzes the critical differences between categorical and numerical variable encoding, and presents compelling evidence of why proper data preprocessing matters for interpretable machine learning. Then render the document to HTML and deploy it via GitHub Pages using the starter repository workflow.

## The Decision Tree Problem ðŸŽ¯

**The Core Problem:** Decision trees are often praised for their interpretability and ability to handle both numerical and categorical variables. But what happens when we encode categorical variables as numbers? How does this affect our understanding of feature importance?

**What is Feature Importance?** In decision trees, feature importance measures how much each variable contributes to reducing impurity (or improving prediction accuracy) across all splits in the tree. It's a key metric for understanding which variables matter most for your predictions.

::: {.callout-important}
## ðŸŽ¯ The Key Insight: Encoding Matters for Interpretability

**The problem:** When we encode categorical variables as numerical values (like 1, 2, 3, 4...), decision trees treat them as if they have a meaningful numerical order. This can completely distort our understanding of feature importance.

**Why this matters:** If a categorical variable like "Zip Code" (50010, 50011, 50012, 50013) is treated as a numerical variable, the tree might split on "Zip Code > 50012.5" instead of recognizing that this is really a categorical choice between discrete geographic areas where the order of numerical values has no meaningful interpretation.

**The connection:** Proper encoding preserves the true nature of categorical variables and gives us accurate feature importance rankings.
:::

**The Devastating Reality:** Even sophisticated machine learning models can give us completely wrong insights about feature importance if we don't properly encode our variables. A categorical variable that should be among the most important might appear irrelevant, while a numerical variable might appear artificially important.

## Mistaken Feature Importance: The Example of Zip Code

Here we load a dataset that has a categorical variable "Zip Code" and a numerical variable "Sale Price" along with a bunch of other truly numerical variables.  In this section, we will explore how the decision tree treats the "Zip Code" when it is mistakenly considered a numerical variable.

### Data Loading and Initial Exploration

::: {.panel-tabset}

### R

```{r}
#| label: load-data-r
#| echo: true
#| message: false
#| warning: false

# Load required libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(rpart))

# Try to load rpart.plot, install if not available
if (!require(rpart.plot, quietly = TRUE)) {
  install.packages("rpart.plot", repos = "https://cran.rstudio.com/")
  library(rpart.plot)
}

# Load the Sales Price dataset
# Note: This loads the data from the buad442Fall2025 repository
sales_data <- read.csv("https://raw.githubusercontent.com/flyaflya/buad442Fall2025/refs/heads/main/datasets/salesPriceData.csv")

# Display basic information about the dataset
cat("Dataset dimensions:", dim(sales_data), "\n")
cat("Number of variables:", ncol(sales_data), "\n")
cat("Number of observations:", nrow(sales_data), "\n\n")

# Show first few rows
head(sales_data, 10)
```

```{r}
#| label: data-structure-r
#| echo: true
#| message: false
#| warning: false

# Examine data types and structure
str(sales_data)

# Summary statistics for key variables
summary(sales_data)
```

### Python

```{python}
#| label: load-data-python
#| echo: true

import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from sklearn.tree import DecisionTreeRegressor, plot_tree
from sklearn.model_selection import train_test_split
from sklearn.metrics import mean_squared_error, r2_score
import warnings
warnings.filterwarnings('ignore')

# Try to import seaborn, but don't fail if it's not available
try:
    import seaborn as sns
    sns.set_style("whitegrid")
except ImportError:
    print("Note: seaborn not available, using matplotlib defaults")

# Load the Sales Price dataset
# Note: This loads the data from the buad442Fall2025 repository
sales_data = pd.read_csv("https://raw.githubusercontent.com/flyaflya/buad442Fall2025/refs/heads/main/datasets/salesPriceData.csv")

print("Dataset loaded successfully!")

# Display basic information about the dataset
print(f"Dataset dimensions: {sales_data.shape}")
print(f"Number of variables: {sales_data.shape[1]}")
print(f"Number of observations: {sales_data.shape[0]}\n")

# Show first few rows
print("First 10 rows:")
sales_data.head(10)
```

```{python}
#| label: data-structure-python
#| echo: true

# Examine data types and structure
print("Data types:")
print(sales_data.dtypes)
print("\nData structure:")
print(sales_data.info())

# Summary statistics for key variables
print("\nSummary statistics for key variables:")
sales_data.describe(include='all')
```
:::

### Dataset Description

The Sales Price dataset contains real estate data with multiple variables describing various aspects of each property. Key variables include:

**Target Variable:**
- `SalePrice`: The sale price of the house (our prediction target)

**Key Variables:**
- `LotArea`: Lot size in square feet
- `YearBuilt`: Year the house was originally built
- `GrLivArea`: Above ground living area square feet
- `FullBath`: Number of full bathrooms
- `HalfBath`: Number of half bathrooms
- `BedroomAbvGr`: Number of bedrooms above grade
- `TotRmsAbvGrd`: Total rooms above grade
- `GarageCars`: Size of garage in car capacity
- `zipCode`: Zip code of the property

## Decision Tree Model Building

Now we'll build a decision tree model to predict house sale prices. We'll treat all predictor variables as numerical (including zipCode) to demonstrate how this affects feature importance interpretation.

### Data Preparation and Model Training

::: {.panel-tabset}

### R

```{r}
#| label: prepare-data-r
#| echo: true
#| message: false
#| warning: false

# Select key variables for the model
# We'll treat zipCode as numerical (which is problematic for interpretation)
model_data <- sales_data %>%
  select(SalePrice, LotArea, YearBuilt, GrLivArea, FullBath, HalfBath, 
         BedroomAbvGr, TotRmsAbvGrd, GarageCars, zipCode) %>%
  na.omit()  # Remove any missing values

# Check for missing values
cat("Missing values check:\n")
sapply(model_data, function(x) sum(is.na(x)))

# Split the data into training and testing sets (80/20 split)
set.seed(123)  # For reproducibility
train_indices <- sample(1:nrow(model_data), 0.8 * nrow(model_data))
train_data <- model_data[train_indices, ]
test_data <- model_data[-train_indices, ]

cat("\nTraining set size:", nrow(train_data), "\n")
cat("Testing set size:", nrow(test_data), "\n")
```

```{r}
#| label: build-tree-r
#| echo: true
#| message: false
#| warning: false

# Build decision tree with maximum depth of 3
# Using rpart for regression tree
tree_model <- rpart(SalePrice ~ ., 
                    data = train_data,
                    method = "anova",
                    control = rpart.control(maxdepth = 3, 
                                          minsplit = 20, 
                                          minbucket = 10))

# Display model summary
cat("Decision Tree Model Summary:\n")
print(tree_model)

# Model complexity parameters
cat("\nModel complexity parameters:\n")
cat("Number of splits:", length(unique(tree_model$where)), "\n")
cat("Number of terminal nodes:", sum(tree_model$frame$var == "<leaf>"), "\n")
```

### Python

```{python}
#| label: prepare-data-python
#| echo: true

# Select key variables for the model
# We'll treat zipCode as numerical (which is problematic for interpretation)
model_vars = ['SalePrice', 'LotArea', 'YearBuilt', 'GrLivArea', 'FullBath', 
              'HalfBath', 'BedroomAbvGr', 'TotRmsAbvGrd', 'GarageCars', 'zipCode']

model_data = sales_data[model_vars].dropna()

# Check for missing values
print("Missing values check:")
print(model_data.isnull().sum())

# Split the data into training and testing sets (80/20 split)
from sklearn.model_selection import train_test_split

X = model_data.drop('SalePrice', axis=1)
y = model_data['SalePrice']

X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.2, random_state=123)

print(f"\nTraining set size: {X_train.shape[0]}")
print(f"Testing set size: {X_test.shape[0]}")
print(f"Number of features: {X_train.shape[1]}")
```

```{python}
#| label: build-tree-python
#| echo: true

# Build decision tree with maximum depth of 3
tree_model = DecisionTreeRegressor(max_depth=3, 
                                  min_samples_split=20, 
                                  min_samples_leaf=10, 
                                  random_state=123)

# Fit the model
tree_model.fit(X_train, y_train)

# Display model summary
print("Decision Tree Model Summary:")
print(f"Number of features: {tree_model.n_features_in_}")
print(f"Number of leaves: {tree_model.get_n_leaves()}")
print(f"Tree depth: {tree_model.get_depth()}")
print(f"Number of nodes: {tree_model.tree_.node_count}")
```

:::

### Tree Visualization

Let's visualize our decision tree to understand how it makes predictions and which variables it considers most important.

::: {.panel-tabset}

### R

```{r}
#| label: visualize-tree-r
#| echo: true
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

# Create a more detailed tree plot
par(mfrow = c(1, 1))

# Try rpart.plot first, fallback to plot if not available
if (require(rpart.plot, quietly = TRUE)) {
  rpart.plot(tree_model, 
             type = 2,  # Show split labels
             extra = 101,  # Show number of observations and percentage
             fallen.leaves = TRUE,  # Position leaf nodes at bottom
             digits = 0,  # Round numbers
             cex = 0.8,  # Text size
             main = "Decision Tree for House Price Prediction\n(Treating zipCode as Numerical)")
} else {
  # Fallback to basic plot
  plot(tree_model, uniform = TRUE, main = "Decision Tree for House Price Prediction\n(Treating zipCode as Numerical)")
  text(tree_model, use.n = TRUE, all = TRUE, cex = 0.8)
}
```

```{r}
#| label: tree-text-r
#| echo: true
#| message: false
#| warning: false

# Print the tree structure in text format
cat("Tree Structure (Text Format):\n")
print(tree_model)
```

### Python

```{python}
#| label: visualize-tree-python
#| echo: true
#| fig-width: 12
#| fig-height: 8

# Create tree visualization
plt.figure(figsize=(12, 8))
plot_tree(tree_model, 
          feature_names=X_train.columns,
          filled=True, 
          rounded=True,
          fontsize=10,
          max_depth=3)
plt.title("Decision Tree for House Price Prediction\n(Treating zipCode as Numerical)")
plt.tight_layout()
plt.show()
```

```{python}
#| label: tree-text-python
#| echo: true

# Print tree structure in text format
from sklearn.tree import export_text

tree_rules = export_text(tree_model, feature_names=list(X_train.columns))
print("Tree Structure (Text Format):")
print(tree_rules)
```

:::

### Model Performance Evaluation

Let's evaluate how well our decision tree performs on both training and testing data.

::: {.panel-tabset}

### R

```{r}
#| label: model-evaluation-r
#| echo: true
#| message: false
#| warning: false

# Make predictions
train_pred <- predict(tree_model, train_data)
test_pred <- predict(tree_model, test_data)

# Calculate performance metrics
train_rmse <- sqrt(mean((train_data$SalePrice - train_pred)^2))
test_rmse <- sqrt(mean((test_data$SalePrice - test_pred)^2))

train_r2 <- 1 - sum((train_data$SalePrice - train_pred)^2) / 
            sum((train_data$SalePrice - mean(train_data$SalePrice))^2)
test_r2 <- 1 - sum((test_data$SalePrice - test_pred)^2) / 
           sum((test_data$SalePrice - mean(test_data$SalePrice))^2)

# Display performance metrics
cat("Model Performance Metrics:\n")
cat("Training RMSE: $", round(train_rmse, 2), "\n")
cat("Testing RMSE: $", round(test_rmse, 2), "\n")
cat("Training RÂ²: ", round(train_r2, 4), "\n")
cat("Testing RÂ²: ", round(test_r2, 4), "\n")

# Calculate mean absolute error
train_mae <- mean(abs(train_data$SalePrice - train_pred))
test_mae <- mean(abs(test_data$SalePrice - test_pred))

cat("Training MAE: $", round(train_mae, 2), "\n")
cat("Testing MAE: $", round(test_mae, 2), "\n")
```

```{r}
#| label: prediction-plot-r
#| echo: true
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Create prediction vs actual plots
par(mfrow = c(1, 2))

# Training data
plot(train_data$SalePrice, train_pred, 
     main = "Training Data: Actual vs Predicted",
     xlab = "Actual Sale Price ($)", 
     ylab = "Predicted Sale Price ($)",
     pch = 16, col = "blue", alpha = 0.6)
abline(0, 1, col = "red", lwd = 2)

# Testing data
plot(test_data$SalePrice, test_pred, 
     main = "Testing Data: Actual vs Predicted",
     xlab = "Actual Sale Price ($)", 
     ylab = "Predicted Sale Price ($)",
     pch = 16, col = "green", alpha = 0.6)
abline(0, 1, col = "red", lwd = 2)

par(mfrow = c(1, 1))
```

### Python

```{python}
#| label: model-evaluation-python
#| echo: true

# Make predictions
train_pred = tree_model.predict(X_train)
test_pred = tree_model.predict(X_test)

# Calculate performance metrics
train_rmse = np.sqrt(mean_squared_error(y_train, train_pred))
test_rmse = np.sqrt(mean_squared_error(y_test, test_pred))

train_r2 = r2_score(y_train, train_pred)
test_r2 = r2_score(y_test, test_pred)

# Display performance metrics
print("Model Performance Metrics:")
print(f"Training RMSE: ${train_rmse:,.2f}")
print(f"Testing RMSE: ${test_rmse:,.2f}")
print(f"Training RÂ²: {train_r2:.4f}")
print(f"Testing RÂ²: {test_r2:.4f}")

# Calculate mean absolute error
from sklearn.metrics import mean_absolute_error

train_mae = mean_absolute_error(y_train, train_pred)
test_mae = mean_absolute_error(y_test, test_pred)

print(f"Training MAE: ${train_mae:,.2f}")
print(f"Testing MAE: ${test_mae:,.2f}")
```

```{python}
#| label: prediction-plot-python
#| echo: true
#| fig-width: 12
#| fig-height: 5

# Create prediction vs actual plots
fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 5))

# Training data
ax1.scatter(y_train, train_pred, alpha=0.6, color='blue')
ax1.plot([y_train.min(), y_train.max()], [y_train.min(), y_train.max()], 'r--', lw=2)
ax1.set_xlabel('Actual Sale Price ($)')
ax1.set_ylabel('Predicted Sale Price ($)')
ax1.set_title('Training Data: Actual vs Predicted')
ax1.grid(True, alpha=0.3)

# Testing data
ax2.scatter(y_test, test_pred, alpha=0.6, color='green')
ax2.plot([y_test.min(), y_test.max()], [y_test.min(), y_test.max()], 'r--', lw=2)
ax2.set_xlabel('Actual Sale Price ($)')
ax2.set_ylabel('Predicted Sale Price ($)')
ax2.set_title('Testing Data: Actual vs Predicted')
ax2.grid(True, alpha=0.3)

plt.tight_layout()
plt.show()
```

:::

## Feature Importance Analysis

Now let's examine which features the decision tree considers most important. This is where we'll see the impact of treating zipCode as a numerical variable.

::: {.panel-tabset}

### R

```{r}
#| label: feature-importance-r
#| echo: true
#| message: false
#| warning: false

# Extract feature importance
importance_df <- data.frame(
  Feature = names(tree_model$variable.importance),
  Importance = as.numeric(tree_model$variable.importance)
) %>%
  arrange(desc(Importance)) %>%
  mutate(Importance_Percent = round(Importance / sum(Importance) * 100, 2))

# Display feature importance
cat("Feature Importance Rankings:\n")
print(importance_df)

# Create a bar plot of feature importance
library(ggplot2)

ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(title = "Feature Importance in Decision Tree Model",
       subtitle = "Variables ranked by their contribution to reducing impurity",
       x = "Features",
       y = "Importance Score") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

```{r}
#| label: importance-analysis-r
#| echo: true
#| message: false
#| warning: false

# Analyze the top features
cat("Top 3 Most Important Features:\n")
top_features <- head(importance_df, 3)
for(i in 1:nrow(top_features)) {
  cat(i, ". ", top_features$Feature[i], 
      " (", top_features$Importance_Percent[i], "% of total importance)\n", sep = "")
}

cat("\nKey Observations:\n")
cat("- zipCode appears as feature #", which(importance_df$Feature == "zipCode"), 
    " with ", importance_df$Importance_Percent[importance_df$Feature == "zipCode"], 
    "% importance\n", sep = "")
cat("- This is problematic because zipCode is being treated as a numerical variable\n")
cat("- The tree might split on 'zipCode > 50012.5' which has no meaningful interpretation\n")
```

### Python

```{python}
#| label: feature-importance-python
#| echo: true

# Extract feature importance
importance_df = pd.DataFrame({
    'Feature': X_train.columns,
    'Importance': tree_model.feature_importances_
}).sort_values('Importance', ascending=False)

importance_df['Importance_Percent'] = (importance_df['Importance'] * 100).round(2)

# Display feature importance
print("Feature Importance Rankings:")
print(importance_df)

# Create a bar plot of feature importance
plt.figure(figsize=(10, 6))
plt.barh(range(len(importance_df)), importance_df['Importance'], 
         color='steelblue', alpha=0.7)
plt.yticks(range(len(importance_df)), importance_df['Feature'])
plt.xlabel('Importance Score')
plt.title('Feature Importance in Decision Tree Model\nVariables ranked by their contribution to reducing impurity')
plt.gca().invert_yaxis()
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()
```

```{python}
#| label: importance-analysis-python
#| echo: true

# Analyze the top features
print("Top 3 Most Important Features:")
top_features = importance_df.head(3)
for i, (_, row) in enumerate(top_features.iterrows(), 1):
    print(f"{i}. {row['Feature']} ({row['Importance_Percent']}% of total importance)")

zipcode_rank = importance_df[importance_df['Feature'] == 'zipCode'].index[0] + 1
zipcode_importance = importance_df[importance_df['Feature'] == 'zipCode']['Importance_Percent'].iloc[0]

print(f"\nKey Observations:")
print(f"- zipCode appears as feature #{zipcode_rank} with {zipcode_importance}% importance")
print("- This is problematic because zipCode is being treated as a numerical variable")
print("- The tree might split on 'zipCode > 50012.5' which has no meaningful interpretation")
```

:::

### Critical Analysis: The Problem with Numerical Encoding

::: {.callout-warning}
## âš ï¸ The Encoding Problem Revealed

**What we just observed:** Our decision tree treated `zipCode` as a numerical variable, allowing it to make splits like "zipCode > 50012.5". This creates several problems:

1. **Meaningless Splits:** A zip code of 50013 is not "greater than" 50012 in any meaningful way for house prices
2. **False Importance:** The algorithm might assign high importance to zipCode simply because it can create many splits
3. **Misleading Interpretations:** We might conclude that zipCode is very important when it's really just an artifact of poor encoding

**The Real Issue:** Zip codes are categorical variables that represent discrete geographic areas. The numerical values (50010, 50011, 50012, etc.) have no inherent order or magnitude relationship to house prices.
:::

**Next Steps:** In the following sections, we'll demonstrate how proper categorical encoding changes both the tree structure and feature importance rankings, revealing the true importance of each variable for predicting house prices.

